---
title: "Perform Principal Component Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{v10_perform_pca}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library("eigentoolbox")
# devtools::load_all()
```


Quick and dirty Eigenanalysis, Part One: **PCA**.


## data
We need a data set. 
Since I first learned this from a text by Prof. Norm MacLeod [1], I will stick with his data
(although, note that there is a comma-dot typo in the `Sphaerexochus` line of the web table, and the mean body length differs, but *so what*).

```{r data}
trilobite_text = "genus,body_length_mm,glabellar_length_mm,glabellar_width_mm
Acaste,23.14,3.50,3.77
Balizoma,14.32,3.97,4.08
Calymene,51.69,10.91,10.72
Ceraurus,21.15,4.90,4.69
Cheirurus,31.74,9.33,12.11
Cybantyx,26.81,11.35,10.10
Cybeloides,25.13,6.39,6.81
Dalmanites,32.93,8.46,6.08
Delphion,21.81,6.92,9.01
Ormathops,13.88,5.03,4.34
Phacopdina,21.43,7.03,6.79
Phacops,27.23,5.30,8.19
Placopoaria,38.15,9.40,8.71
Pricyclopyge,40.11,14.98,12.98
Ptychoparia,62.17,12.25,8.71
Rhenops,55.94,19.00,13.10
Sphaerexochus,23.31,3.84,4.60
Toxochasmops,46.12,8.15,11.42
Trimerus,89.43,23.18,21.52
Zacanthoides,47.89,13.56,11.78
"

data <- read.table(
    text = trilobite_text,
    sep = ",",
    header = TRUE,
    stringsAsFactors = FALSE
)

summary(data)
```

## Calculation

First, perform the SVD of the covariance matrix.

```{r pca}
pca <- compute_pca(data)
summary(data)
```


Does this PCA match the referenced one?
```{r}
reference <- read.table(
    text = "1,383.08, 0.951,0.244,0.192
            2,  7.46, 0.310,0.700,0.644
            3,  2.00,−0.023,0.671,0.741",
    sep = ",",
    header = FALSE,
    stringsAsFactors = FALSE
)
colnames(reference) <- c("i", "ev", "l1", "l2", "l3")
knitr::kable(reference)
```


```{r pca_matrix_match}
test <- cbind(1:3, pca$val, t(pca$mat))
colnames(test) <- c("i", "ev", "l1", "l2", "l3")
knitr::kable(test)
```

Well, ... approximately.



## Forward Trafo

No analysis without data - so transform the data.

```{r transform}
data_trafo <- transform( data, pca )

summary(data_trafo)
```

## Inverse Trafo

Transformation can be undone:

```{r retransform}
data_retrafo <- retrafo( data_trafo, pca )

summary(data_retrafo)
```


Did we correctly invert?

```{r}
sum(data[, pca$features] - data_retrafo)
```


This is clearly the "minimum expected, sufficient success" (MESS™).



## References

- [1] Norm MacLeod (2005): "PalaeoMath: Part 5 - PCA, Eigenanalysis & Regression V", Palaeontology Newsletter, edition Nº 59. https://www.palass.org/publications/newsletter/palaeomath-101/palaeomath-part-5-pca-eigenanalysis-regression-v
